Do Alternate Wetting and Drying Irrigation Technology and Nitrogen Rates Affect Rice Sheath Blight?

### 2018-11-01 analysis/paper/paper.Rmd
Replace spaces with non-breaking

Sheath blight disease (*Rhizoctonia solani* Kühn), anastomosis group 1 [(teleomorph: *Thanatephorus cucumeris* (A.B. Frank) Donk.] of rice is an economically important disease worldwide throughout tropical and temperate production areas. Under conducive conditions the disease can cause up to 50 \% yield losses [@Marchetti1991]. Studies by Savary et al. [-@Savary2000] estimated normal yield losses in tropical irrigated lowland rice in Asia at 5-10 \%. The disease is best managed through an integrated disease management approach which includes resistance [@Kumar2009; @Srinivasachary2011] fungicide applications where affordable [@Rush1983; @Groth2008], nitrogen (N) management [@Slaton2003; @Castilla1996; @Tang2007] and water management [@Castilla1996].

Alternate wetting and drying (AWD) is an irrigation technique for rice (*Oryza sativa* L.) developed by the International Rice Research Institute (IRRI) and its partners that saves about 15-40 \% of irrigation water while not affecting yield [@Bouman2001; @Feng2007]. In AWD rice, fields are exposed to several dry phases during the growing season without exposing the plants to water stress. In order to avoid yield decline under AWD “safe” thresholds have been developed. Under safe AWD irrigation water is applied when the field water level reaches 15cm below the soil surface [@Richards2014]. Fields are furthermore kept flooded during the flowering period to avoid plant stress. As an added benefit to saving water, AWD also reduces greenhouse gas (GHG) emissions of rice fields, which are a substantial factor in the GHG budget of rice producing countries, by around 50 \% [@Yan2005; @Sander2015].

@@

The split-plot treatment rates of N were N0 (no N supply), N100 (100 kg per ha applied as urea in three splits) and N120 (120 kg per ha applied as urea in three splits) (Tab. \@ref(tab:tab-1)).

@@

Split-plot N rates differed from the 2015 rates with only two N treatments, N60 (60 kg per ha as urea split into two applications) and N180 (180 kg per ha in three splits) being applied (Tab. \@ref(tab:tab-1)).

@@

In both experiments, two sample areas, sized 1m x 1m, per plot were assessed. The total number of tillers per hill and number of tillers with sheath blight (incidence) were measured for nine hills per sample area. Tiller sheath severity was measured for four tillers per hill and six leaves tiller using a disease assessment rating scale for field evaluation developed at IRRI. The scale is an unequally spaced categorical scale where 0 - no disease; 1 - trace to 1 \% severity; 2 - 1 to 5 \%; 3 - 5 to 15 \%; 4 - 15 to 50 \%; 5 - 50 to 100 \%. Severity was considered to be the amount of leaf or sheath tissue covered by sheath blight lesions. Leaves were rated as green (living) having at least XXX \% living tissue or dead, defined as having less than XXX \% living tissue. The same rating methodology was used in both experiments. Five disease assessments were made in the 2015 experiment and four disease assessments were made in the 2016 experiment, respectively.

@@

In 2015 the incidence of tiller sheath blight remained low throughout the growing season (Fig. \@ref(fig:fig-1)a, \@ref(fig:fig-2)a). Water management was not statistically clearly different (Fig. \@ref(fig:fig-3)a, \@ref(fig:fig-5)a). However, the N treatments, N100 and N120, were both statistically clearly different when compared with the control N0 treatment (Fig. \@ref(fig:fig-4)a, \@ref(fig:fig-5)a).

@@

Tiller sheath blight severity remained below 2% (Fig. \@ref(fig:fig-2)c, \@ref(fig:fig-1)c). Both the N100 and N120 treatments were statistically clearly different than the control N0 treatment (Fig. \@ref(fig:fig-4)c, \@ref(fig:fig-5)c). However, the AWD water management was not statistically clearly different from the flooded treatment (Fig. \@ref(fig:fig-3)c, \@ref(fig:fig-5)c).

@@

Severity of leaf sheath blight remained low, less than 0.4% across all treatments (Fig. \@ref(fig:fig-1)e, \@ref(fig:fig-2)e). Both the N100 and N120 treatments were statistically clearly different than the N0 treatment (Fig. \@ref(fig:fig-4)e, \@ref(fig:fig-5)e). However, the AWD water management was not statistically clearly different from the flooded treatment (Fig. \@ref(fig:fig-3)e, \@ref(fig:fig-5)e).

@@

The changes to the inoculation methodology resulted in a higher rate of infection in 2016, with the N180 treatment reaching a maximum value of 98% incidence at the third observation (Fig. \@ref(fig:fig-2)b). The N treatment N180, was statistically clearly different than the control N60 treatment (Fig. \@ref(fig:fig-4)b, \@ref(fig:fig-5)b). As in the 2015 study, water management did not statistically clearly differ (Fig. \@ref(fig:fig-3)b, \@ref(fig:fig-5)b).

@@

As with the tiller incidence, the tiller severity increased with the changed inoculation methods with a maximum of 7.6% for the flooded treatment (Fig. \@ref(fig:fig-2)d). In 2016 the N180 treatment was statistically clearly different from the N60 treatment, with N180 severity being higher (Fig. \@ref(fig:fig-4)d, \@ref(fig:fig-5)d). The AWD water management, which was lower than the flooded treatment, was also statistically clearly different (Fig. \@ref(fig:fig-3)d, \@ref(fig:fig-5)d).

@@

In 2016 the neither of the treatments, N rate or water management, were statistically clearly different from the control treatments for leaf sheath blight severity (Fig. \@ref(fig:fig-3)f, \@ref(fig:fig-5)f, \@ref(fig:fig-4)f, \@ref(fig:fig-5)f).

@@

In both experiments we were unable to detect any statistically clear effect of AWD on sheath blight that led to increases in the disease that could hinder adoption of the technology. In fact, in only one instance were we able to detect any clear effect of AWD on sheath blight. In the 2016 experiment the tiller sheath blight severity was clearly lower for the AWD treatment than flooded treatment (Fig. \@ref(fig:fig-3)d, \@ref(fig:fig-5)d), indicating a possible adverse effect of using AWD on tiller sheath blight severity under high sheath blight pressure.

@@

By increasing the plot size and increasing inoculum amount applied to a smaller area, the changes made for the 2016 experiment appear to have improved the experiment. The the sheath blight incidence increased and the variability of sheath blight in the plots decreased (Fig. \@ref(fig:fig-1)a:b, \@ref(fig:fig-2)c:d, \@ref(fig:fig-3)a:b, \@ref(fig:fig-4)c:d).

@@

```{r fig-5, fig.cap="Posterior means and 95% credible intervals for the explanatory variables used in models of alternate wetting and drying irrigations (AWD) and N rate (NRTE) on sheath blight tiller incidence and tiller severity. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180)."}

@@

```{r fig-6, fig.cap="Posterior means and 95% credible intervals for the explanatory variables used in models of alternate wetting and drying irrigations (AWD) and N rate (NRTE) and the interaction of the two treatments on sheath blight leaf severity. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180)."}

@@

```{r colophon, cache = FALSE}
# which R packages and versions?
devtools::session_info()
```

>>>>

Sheath blight disease (*Rhizoctonia solani* Kühn), anastomosis group 1 [(teleomorph: *Thanatephorus cucumeris* (A.B. Frank) Donk.] of rice is an economically important disease worldwide throughout tropical and temperate production areas. Under conducive conditions the disease can cause up to 50&nbsp;\% yield losses [@Marchetti1991]. Studies by Savary et al. [-@Savary2000] estimated normal yield losses in tropical irrigated lowland rice in Asia at 5-10 &nbsp;\%. The disease is best managed through an integrated disease management approach which includes resistance [@Kumar2009; @Srinivasachary2011] fungicide applications where affordable [@Rush1983; @Groth2008], nitrogen (N) management [@Slaton2003; @Castilla1996; @Tang2007] and water management [@Castilla1996].

Alternate wetting and drying (AWD) is an irrigation technique for rice (*Oryza sativa* L.) developed by the International Rice Research Institute (IRRI) and its partners that saves about 15-40&nbsp;\% of irrigation water while not affecting yield [@Bouman2001; @Feng2007]. In AWD rice, fields are exposed to several dry phases during the growing season without exposing the plants to water stress. In order to avoid yield decline under AWD “safe” thresholds have been developed. Under safe AWD irrigation water is applied when the field water level reaches 15cm below the soil surface [@Richards2014]. Fields are furthermore kept flooded during the flowering period to avoid plant stress. As an added benefit to saving water, AWD also reduces greenhouse gas (GHG) emissions of rice fields, which are a substantial factor in the GHG budget of rice producing countries, by around 50 &nbsp;\% [@Yan2005; @Sander2015].

@@

The split-plot treatment rates of N were N0 (no N supply), N100 (100 kg per ha applied as urea in three splits) and N120 (120 kg per ha applied as urea in three splits) (Tab.&nbsp;\@ref(tab:tab-1)).

@@

Split-plot N rates differed from the 2015 rates with only two N treatments, N60 (60 kg per ha as urea split into two applications) and N180 (180 kg per ha in three splits) being applied (Tab.&nbsp;\@ref(tab:tab-1)).

@@

In both experiments, two sample areas, sized 1m x 1m, per plot were assessed. The total number of tillers per hill and number of tillers with sheath blight (incidence) were measured for nine hills per sample area. Tiller sheath severity was measured for four tillers per hill and six leaves tiller using a disease assessment rating scale for field evaluation developed at IRRI. The scale is an unequally spaced categorical scale where 0 - no disease; 1 - trace to 1&nbsp;\% severity; 2 - 1 to 5&nbsp;\%; 3 - 5 to 15&nbsp;\%; 4 - 15 to 50&nbsp;\%; 5 - 50 to 100&nbsp;\%. Severity was considered to be the amount of leaf or sheath tissue covered by sheath blight lesions. Leaves were rated as green (living) having at least XXX&nbsp;\% living tissue or dead, defined as having less than XXX&nbsp;\% living tissue. The same rating methodology was used in both experiments. Five disease assessments were made in the 2015 experiment and four disease assessments were made in the 2016 experiment, respectively.

@@

In 2015 the incidence of tiller sheath blight remained low throughout the growing season (Fig.&nbsp;\@ref(fig:fig-1)a, \@ref(fig:fig-2)a). Water management was not statistically clearly different (Fig.&nbsp;\@ref(fig:fig-3)a, \@ref(fig:fig-5)a). However, the N treatments, N100 and N120, were both statistically clearly different when compared with the control N0 treatment (Fig.&nbsp;\@ref(fig:fig-4)a, \@ref(fig:fig-5)a).

@@

Tiller sheath blight severity remained below 2&nbsp;\% (Fig.&nbsp;\@ref(fig:fig-2)c, \@ref(fig:fig-1)c). Both the N100 and N120 treatments were statistically clearly different than the control N0 treatment (Fig.&nbsp;\@ref(fig:fig-4)c, \@ref(fig:fig-5)c). However, the AWD water management was not statistically clearly different from the flooded treatment (Fig.&nbsp;\@ref(fig:fig-3)c, \@ref(fig:fig-5)c).

@@

Severity of leaf sheath blight remained low, less than 0.4&nbsp;% across all treatments (Fig.&nbsp;\@ref(fig:fig-1)e, \@ref(fig:fig-2)e). Both the N100 and N120 treatments were statistically clearly different than the N0 treatment (Fig.&nbsp;\@ref(fig:fig-4)e, \@ref(fig:fig-5)e). However, the AWD water management was not statistically clearly different from the flooded treatment (Fig.&nbsp;\@ref(fig:fig-3)e, \@ref(fig:fig-5)e).

@@

The changes to the inoculation methodology resulted in a higher rate of infection in 2016, with the N180 treatment reaching a maximum value of 98&nbsp;% incidence at the third observation (Fig.&nbsp;\@ref(fig:fig-2)b). The N treatment N180, was statistically clearly different than the control N60 treatment (Fig.&nbsp;\@ref(fig:fig-4)b, \@ref(fig:fig-5)b). As in the 2015 study, water management did not statistically clearly differ (Fig.&nbsp;\@ref(fig:fig-3)b, \@ref(fig:fig-5)b).

@@

As with the tiller incidence, the tiller severity increased with the changed inoculation methods with a maximum of 7.6% for the flooded treatment (Fig.&nbsp;\@ref(fig:fig-2)d). In 2016 the N180 treatment was statistically clearly different from the N60 treatment, with N180 severity being higher (Fig.&nbsp;\@ref(fig:fig-4)d, \@ref(fig:fig-5)d). The AWD water management, which was lower than the flooded treatment, was also statistically clearly different (Fig.&nbsp;\@ref(fig:fig-3)d, \@ref(fig:fig-5)d).

@@

In 2016 the neither of the treatments, N rate or water management, were statistically clearly different from the control treatments for leaf sheath blight severity (Fig.&nbsp;\@ref(fig:fig-3)f, \@ref(fig:fig-5)f, \@ref(fig:fig-4)f, \@ref(fig:fig-5)f).

@@

In both experiments we were unable to detect any statistically clear effect of AWD on sheath blight that led to increases in the disease that could hinder adoption of the technology. In fact, in only one instance were we able to detect any clear effect of AWD on sheath blight. In the 2016 experiment the tiller sheath blight severity was clearly lower for the AWD treatment than flooded treatment (Fig.&nbsp;\@ref(fig:fig-3)d, \@ref(fig:fig-5)d), indicating a possible adverse effect of using AWD on tiller sheath blight severity under high sheath blight pressure.

@@

By increasing the plot size and increasing inoculum amount applied to a smaller area, the changes made for the 2016 experiment appear to have improved the experiment. The the sheath blight incidence increased and the variability of sheath blight in the plots decreased (Fig.&nbsp;\@ref(fig:fig-1)a:b, \@ref(fig:fig-2)c:d, \@ref(fig:fig-3)a:b, \@ref(fig:fig-4)c:d).

@@

```{r fig-5, fig.cap="Posterior means and 95&nbsp;% credible intervals for the explanatory variables used in models of alternate wetting and drying irrigations (AWD) and N rate (NRTE) on sheath blight tiller incidence and tiller severity. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180)."}

@@

```{r fig-6, fig.cap="Posterior means and 95&nbsp;% credible intervals for the explanatory variables used in models of alternate wetting and drying irrigations (AWD) and N rate (NRTE) and the interaction of the two treatments on sheath blight leaf severity. Five different N treatments were applied to split plots depending on the experiment. In the 2015 experiment three N rate treatments were applied: no N (N0), 100 kg per ha N (N100) and 180 kg per ha N (N120). In the 2016 experiment, two N rate treatments were applied: 60 kg per ha N (N60) and 180 kg per ha N (N180)."}

@@

```{r colophon, cache = FALSE}
# which R packages and versions?
sessioninfo::session_info()
```
